\thispagestyle{plain}
\begin{center}
    \vspace{0.8cm}
    \rule{18cm}{0.5pt}
    \raggedright
    
    \huge

    \textbf{Acknowledgements}\\
    \rule{18cm}{0.5pt}
    
    \vspace{1.4cm}
    
    \normalsize
    
    I would first like to thank my wonderful supervisors, Gavin Huttley and Benjamin Kaehler. 
    
    \vspace{0.4cm}
    
    Gavin, I cannot overstate my gratitude for you. The level of care and guidance you have given is remarkable, I feel so lucky to have been your student. Thank you for always pushing me to truly understand why I am doing something, and for never letting me write sentences longer than 30 words. You have helped me grow so much this year and I will be forever grateful for your mentorship. 
    
    \vspace{0.4cm}
    
    Ben, my eigen'visor. I am so grateful for your supervision. Your support, knowledge and encouragement was a truly wonderful thing to have on my side. Your ability to bring humour into (literally) any situation have livened up many a lab meeting, and brightened up this year. 
    
    \vspace{0.4cm}    
    
        
    Thank you to Le Anh-Phuong. Phuong, my labmate, my Honours twin. You have helped me so much this year, with LaTeX code, seminars, colour palettes, oh the list goes on. Your remarkable resilience this year has been inspiring, you are a force to be reckoned with. Hope you have a big nap when this is all over!
    
    \vspace{0.4cm}  
    
    Thank you to Cheng Soon Ong, your never-ending supply of meta questions forced me to think critically and always helped me see parts of my project in a different light.
    
    \vspace{0.4cm}  
    
    Thank you to Eric Stone, whose statistical insights were of such value to this project. 

    \vspace{0.4cm}  
    
    I would like to thank my housemates, Isobel and Viv. Iz, never in our many years of best-friendship have we attempted something as ambitious as doing Honours together and honestly, I don't know how this year would have been possible without you by my side. Thank you for everything. Viv, thank you for your tender care and support, and for putting up with \textit{all that biology nonsense} at the table.  Oh and of course, thanks to Quince for being the emotional rock of the house. 
    
    \vspace{0.4cm}  
    
    Thanks to the rest of the Honours cohort, and to the wider RSB, especially those who partook in day time \textit{and} night time beverages with me. The sense of community you all gave this year was delightful. 
    
    \vspace{0.4cm}  

    Thanks to Kelsey for always restoring my battered love of statistics with those precious ayto viewings. 
    
    \vspace{0.4cm}  
    
     My thanks go to everyone who read a draft, Kate, Soraya, Dad, Von Bing and Jess (a lil' squish with a big stamina for proof reading). 
    
    \vspace{0.4cm}  
    
    Thank you Mum and Dad, thank you for your support, and for always sending me home from family dinner with some fresh eggs and something from the garden. 
   
    \vspace{0.4cm}  
    
    And finally, thank you to my examiners, Jana, Marcin, and Justin. Thank you for questions, encouragement and kindness. It was truly lovely to have you as my panel. 


\end{center}